---
title: "NCYC Energy models"
author: "Riley M. Anderson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
graphics: yes
output:
  github_document:
    toc: yes
    toc_depth: 5
    pandoc_args: --webtex
  html_document:
    keep_md: yes
    theme: readable
    mathjax: default
  html_notebook:
    code_folding: hide
    theme: readable
    mathjax: default
  pdf_document:
    toc: yes
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
---

```{r setup, include = F}
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Knitr Options
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Set root directory to the project directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())


# Set default knitr options: 
# Suppress warnings and messages, cache chunks, 
# Set default figure size to 6x8 at 300 dpi, and save a png and pdf
knitr::opts_chunk$set(warning = F, message = F, collapse = T, cache = T,
    fig.height = 6, fig.width = 8, dpi = 300, # 6x8" @ 300dpi:1800x2400=4.3MP
    dev = c('png', 'pdf'), dev.args = list(pdf = list(onefile = F)))

```



## Overview

This markdown shows the results from 5 BEopt (Building Energy Optimization) models. BEopt takes inputs describing a building, everything from azimuth to water heater set point and simulates seasonal weather patterns for the nearest locale (Hartford, CT) to determine energy usage.

Five models were built to describe:

1) the current clubhouse structure as is (March 2026)

2) insulate south wall with R15 mineral wool and replace the windows with modern, double pane, insulated, low-e windows

3) condition 2 plus an awning on the south wall that completely shades the windows in the summer months

4) condition 3 plus air sealing the ceiling with drywall and insulating the attic to R60 (cellulose or batts)

5) condition 4 plus replacing every light with LEDs

**Assumptions made**

* Framing is 2x4, 16" o.c. with marginal insulation, ~ R7 (old fiberglass batts)
* Current air leakage is substantial 10-15 ACH50
* Most air leakage occurs at the ceiling-attic interface
* Attic insulation is sub optimal (both in amount and installation) ~ R20
* Current windows are a mixture of single and double glazed, uninsulated, leaky ~ U value: > 0.6
* Airtight ceiling drywall brings total air leakage down to 5 ACH50
* All current lighting is not LED
* Plug loads are 25% of normal household usage
* Cooling set point is 76 F with conditioning between June 1 and Sep 1
* BEopt doesn't directly model seasonally-occupied buildings so inputs schedules were adjusted to approximate seasonal usage from May 1 to Nov 15
* Water heater and plumbing components were excluded from these models to simplify energy schedules and limit the results to electric consumers only (our water heater is propane)


### Summary of Results
* BEopt cannot tell us how much cooler the clubhouse will be with these upgrades but it can tell us the *cooling loads* across each condition. We'll use these loads as proxies for the reduction in heat in the summer.

* In its current condition, the clubhouse would use about 1 MWh for cooling.

* Insulating the south wall and replacing the windows would cut that amount by 25%. Continuing with the energy upgrades, simply adding an awning that shades the south facing windows would cut cooling loads by 47%. With the aforementioned upgrades and properly insulating the attic and air sealing the ceiling would cut the cooling loads down by 69%.


* Across the five models, **lighting** has the biggest impact on total energy usage, so if it hasn't been done already, converting every light to LED should be a top priority.


## BEopt model results:

```{r Main_Code, include = F, cache = F}

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Setup - This code is run, but output is hidden
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Load Packages
library(tidyverse) # Needed for data wrangling: dplyr, tidyr, ggplot2
library(cowplot)


# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Data Preparation
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Import datasets

dat <- read.csv("data/be_opt_output.csv")


```


```{r Data_Wrangling, echo = F, comment = ""}

dat1 <- dat %>% 
  pivot_longer(cols = c(current : opt4),
               names_to = "options") %>% 
  mutate(across(c(item : options), factor),
         value = case_when(
           item == "Misc." | item == "Lights" | item == "Stove_Fridge" ~
             round(value * (7/12),3),
           TRUE ~ value
         ),
         kWh = round(value, 4),
         MWh = round(kWh / 1000, 2))

```


```{r be_opt_results, echo = F}

dat1 %>%
  filter(item != "Cooling Fan/Pump",
         item != "Heating") %>% 
  ggplot(aes(x = options, y = MWh, fill = item)) +
  geom_col() +
  geom_label(aes(x = options, y = ,
                 label = MWh),
             position = position_stack(vjust = 0.5),
             show.legend = F, size = 3) +
  scale_fill_manual(values = c("turquoise",
                               "yellow3",
                               "wheat2",
                               "tan2"),
                    labels = c("Cooling",
                              "Lighting",
                              "Misc. & outlets",
                              "Appliances")) +
  theme_minimal(base_size = 14) +
  scale_x_discrete(labels = c("Current",
                              "R15 walls &\nnew windows",
                              "+ South awning",
                              "+ R60 attic &\nceiling drywall",
                              "+ 100% LED\nlights")) +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 0.8)) +
  labs(x = NULL, fill = NULL)

```


```{r cooling_loads, echo = F}

perc_cooling <- dat1 %>% 
  group_by(options) %>% 
  summarise(total = sum(MWh)) %>% 
  left_join(., dat1, by = "options") %>% 
  filter(item == "Cooling") %>% 
  mutate(prop_mwh = MWh / total,
         cost = kWh * 0.33) %>% 
  filter(options != "opt4")

percent_red <- c(1 - perc_cooling$prop_mwh[1] / perc_cooling$prop_mwh[1],
                 1 - perc_cooling$prop_mwh[2] / perc_cooling$prop_mwh[1],
                 1 - perc_cooling$prop_mwh[3] / perc_cooling$prop_mwh[1],
                 1 - perc_cooling$prop_mwh[4] / perc_cooling$prop_mwh[1])


perc_cooling$percent_red <- percent_red


```

## Cooling loads

```{r cooling_figs, echo = F, fig.width=10, fig.height=6}

A <- perc_cooling %>% 
  ggplot(aes(x = options, y = kWh)) +
  geom_col(fill = "turquoise") +
  geom_label(aes(x = options, y = kWh,
                 label = kWh),
             color = "grey45") +
  scale_x_discrete(labels = c("Current",
                              "R15 walls &\nnew windows",
                              "+ South awning",
                              "+ R60 attic &\nceiling drywall")) +
  labs(x = NULL) +
  theme_minimal(base_size = 14) +
  geom_label(aes(x = options, y = kWh,
                 label = paste("$", cost, sep = "")),
             vjust = 2) +
  theme(axis.text.x = element_text(angle = 30,
                                   hjust = 0.8))


B <- perc_cooling %>% 
  ggplot(aes(x = options, y = prop_mwh)) +
  geom_col(fill = "coral2") +
  geom_label(aes(x = options, y = prop_mwh,
                 label = round(prop_mwh, 2)),
             color = "grey45") +
  scale_x_discrete(labels = c("Current",
                              "R15 walls &\nnew windows",
                              "+ South awning",
                              "+ R60 attic &\nceiling drywall")) +
  labs(x = NULL, y = "Proportion of electric bill") +
  theme_minimal(base_size = 14) +
  geom_label(aes(x = options, y = prop_mwh,
                 label = paste(round(-percent_red * 100, 0),
                               "%", sep = "")),
             vjust = 2) +
  theme(axis.text.x = element_text(angle = 30,
                                   hjust = 0.8))



plot_grid(A, B, nrow = 1)


```


## Session Information

```{r Session_Info, echo = F, comment = ""}

# Add session information to help with reproduceability
sessionInfo()


```


