---
title: "Heat Advisory Days since 1980 at NCYC"
author: "Riley M. Anderson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
graphics: yes
output:
  github_document:
    toc: yes
    toc_depth: 5
    pandoc_args: --webtex
  html_document:
    keep_md: yes
    theme: readable
    mathjax: default
  html_notebook:
    code_folding: hide
    theme: readable
    mathjax: default
  pdf_document:
    toc: yes
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
---

```{r setup, include = F}
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Knitr Options
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Set root directory to the project directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())


# Set default knitr options: 
# Suppress warnings and messages, cache chunks, 
# Set default figure size to 6x8 at 300 dpi, and save a png and pdf
knitr::opts_chunk$set(warning = F, message = F, collapse = T, cache = T,
    fig.height = 6, fig.width = 8, dpi = 300, # 6x8" @ 300dpi:1800x2400=4.3MP
    dev = c('png', 'pdf'), dev.args = list(pdf = list(onefile = F)))

```



## Overview

The analysis pulls daily weather data from [DAYMET](https://daymet.ornl.gov/) (ORNL & NASA) to provide estimates of gridded temperature and vapor pressure surfaces for the **NCYC location since 1980**. The objective is to show how many heat advisory days (NWS criteria) have occurred each year. Heat index was calculated with the Rothfusz regression [(Rothfusz 1990)](https://www.wpc.ncep.noaa.gov/html/heatindex_equation.shtml).

Throughout the summer, the **NCYC clubhouse regularly reaches temperatures above 90°F**. This not only makes for an uncomfortable and unwelcoming environment, but it poses a major health risk to those over 60 years of age and young children [(Kenny et al. 2010)](https://www.cmaj.ca/content/182/10/1053). That is, precisely the population using the clubhouse at these times.



### Summary of Results
* 38 of the last 45 years have had at least one day where a heat advisory would be issued by the National Weather Service.

* The number of heat advisory days/year since 1980 ranged from 0-10 with mean 3.8

* Over the last 45 years, between June 1 and September 15, feels-like temperatures at the clubhouse were above 85°F `r round(1244/4815, 2)*100`% of the time. `r round(509/4815, 2) *100`% of this time was above 90°F and 171 heat advisories (>95°F) were issued (`r round(171/4815, 2)*100`%).

* Serious efforts to cool the clubhouse should be considered.

```{r Main_Code, include = F, cache = F}

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Setup - This code is run, but output is hidden
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Load Packages
library(tidyverse) # Needed for data wrangling: dplyr, tidyr, ggplot2
library(cowplot) # Needed for publication-quality ggplots
library(daymetr)
library(mgcv)

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Data Preparation
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

lat <- 41.289640

lon <- -72.367728


##############################
# Functions
heat_index_daymet <- function(tmax, vp){
  Tm <- tmax * 9/5 + 32
  es <- 610.8 * exp((17.27 * tmax) / (tmax + 237.3))
  RH <- (vp / es) * 100
    
  HI <- -42.379 + 2.04901523 * Tm + 10.14333127 * RH - 
      0.22475541 * Tm * RH - 0.00683783 * Tm^2 - 
      0.05481717 * RH^2 + 0.00122874 * Tm^2 * RH +
      0.00085282 * Tm * RH^2 - 0.00000199 * Tm^2 * RH^2
    
  HI[Tm < 80 | RH < 40] <- NA
  HI

}


heat_index_fun <- function(Temp, RH){
    HI <- -42.379 + 2.04901523 * Temp + 10.14333127 * RH - 
      0.22475541 * Temp * RH - 0.00683783 * Temp^2 - 
      0.05481717 * RH^2 + 0.00122874 * Temp^2 * RH +
      0.00085282 * Temp * RH^2 - 0.00000199 * Temp^2 * RH^2
    
    HI
}

rel_hum_fun <- function(tmax, vp){
  es <- 610.8 * exp((17.27 * tmax) / (tmax + 237.3))
  RH <- (vp / es) * 100
  RH
}
```


```{r Data_Wrangling, echo = F, comment = ""}

daymet_data <- download_daymet(
  site = "Old_Saybrook_CT",
  lat = lat,
  lon = lon,
  start = 1980,
  end = as.numeric(format(Sys.Date(), "%Y")) - 1,
  internal = TRUE,
  simplify = TRUE
)

dat1 <- daymet_data %>% 
  pivot_wider(names_from = measurement,
              values_from = value) %>% 
  mutate(date = as.Date(paste(year, yday, sep = "-"), "%Y-%j")) %>% 
  arrange(date) %>% 
  rename(dayl = dayl..s.,
         prcp = prcp..mm.day.,
         tmax = tmax..deg.c.,
         tmin = tmin..deg.c.,
         vp = vp..Pa.)

dat1 <- dat1 %>% 
  mutate(heat_index = heat_index_daymet(tmax, vp),
         tmax_f = tmax * 9/5 + 32,
         RH = rel_hum_fun(tmax, vp))



```


### Heat advisory days at NCYC since 1980
```{r days_above_90, echo = F}

dat1 %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarise(HAD = sum(heat_index > 95, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = HAD)) +
  geom_point(shape = 21) +
  geom_smooth(method = "gam",
              formula = y ~ s(x, bs = "tp"),
              alpha = 0.2) +
  theme_minimal(base_size = 14) +
  labs(x = "Year", y = "Heat advisory days")


```

* **Heat advisory days** per year since 1980 at the NCYC clubhouse (41.289, -72.368). The national weather service defines heat advisory days when the heat index (feels-like temperature) is at or above 95°F for 2 hours or more. Since the data source, DAYMET only provides Tmax on a daily basis (not hourly), the data shown may be slightly overestimated. The best fit hindcast (blue line) was generated with a generalized additive model with thin plate splines and should not be construed as a predictive (forecast) model.


```{r heat_index_figure, echo = F}

top_10th_days <- dat1 %>% 
  select(year, heat_index, tmax_f, RH) %>% 
  filter(heat_index > quantile(heat_index, 0.9,
                               na.rm = T))



temps <- seq(80, 105, by = 0.5)
RHs <- seq(20, 100, by = 1)


temp_grid <- expand.grid(temp = temps,
                         RH = RHs)

temp_grid <- temp_grid %>% 
  mutate(heat_index = heat_index_fun(temp, RH))

# temp_grid %>% 
#   ggplot(aes(x = temp, y = RH)) +
#   geom_contour_filled(aes(z = heat_index),
#                       breaks = c(-Inf, 95, Inf),
#                       alpha = 0.5) +
#   geom_contour(aes(z = heat_index),
#                breaks = 95,
#                color = "black",
#                linewidth = 1) +
#   labs(x = "Air temperature (°F)",
#        y = "Relative humidity (%)",
#        fill = "Heat-index zone") +
#   theme_minimal() +
#   geom_point(data = top_10th_days,
#              aes(x = tmax_f, y = RH, colour = year)) +
#   scale_color_viridis_c(option = "A", direction = -1) +
#   scale_fill_manual(values = c("grey95", "grey65"),
#                     labels = c("<95F", ">95F"))

```


### How hot is it at the clubhouse?

```{r density_HAD_fig, echo = F}


temp_grid %>% 
  ggplot(aes(x = temp, y = RH)) +
  geom_contour(aes(z = heat_index),
               breaks = c(85, 90, 95),
               linewidth = 1.2,
               color = "black") +
  stat_density_2d(data = filter(dat1,
                                yday > 150 & yday < 250),
                  aes(tmax_f, RH),
                  color = "blue") +
  labs(x = "Air temperature (°F)",
       y = "Relative humidity (%)",
       fill = "Feels-like temp") +
  theme_minimal(base_size = 20) +
  geom_contour_filled(aes(z = heat_index),
                      breaks = c(-Inf, 95, Inf),
                      alpha = 0.3) +
  geom_contour_filled(aes(z = heat_index),
                      breaks = c(90, 95),
                      alpha = 0.5) + 
  geom_contour_filled(aes(z = heat_index),
                      breaks = c(85, 90),
                      alpha = 0.5) +
  scale_fill_manual(values = c("white", "firebrick1",
                               "orange1", "yellow"),
                    labels = c("", "Heat Advisory\n>95F", ">90F", ">85F"))


```

* **Density contours of the daily maximum temperature from June 1 to Sep 15.** Blue contours show the density of the raw data with frequency diminishing outwards. Black isoclines overlay the heat index where yellow indicates feels-like temperatures between 85°F and 90°F, orange is 90°F to 95°F, and pink is >95°F. The national weather service issues heat advisories when the heat index is 95°F or greater for 2 hours or more. The NCYC clubhouse regularly experiences feels-like temperatures above 85°F



```{r GAM, echo = F, eval = F}


dat2 <- dat1 %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarise(HAD = sum(heat_index > 95, na.rm = TRUE))


gam_fit1 <- gam(HAD ~ s(year, k = 10),
                data = dat2,
                family = poisson())

summary(gam_fit1)
# gam.check(gam_fit1)
# plot(gam_fit1, shade = T)


future_years <- data.frame(year = seq(max(dat2$year),
                                      max(dat2$year) + 5))

pred <- predict(gam_fit1,
                newdata = future_years,
                type = "link",
                se.fit = TRUE)

future_years <- future_years %>% 
  mutate(fit = exp(pred$fit),
         lwr = exp(pred$fit - 2 * pred$se.fit),
         upr = exp(pred$fit + 2 * pred$se.fit))

```

```{r forecast_HADs, echo = F, eval = F}

dat2 %>% 
  ggplot(aes(x = year, y = HAD)) +
  geom_point() +
  geom_smooth(method = "gam",
              formula = y ~ s(x, bs = "tp")) +
  geom_line(data = future_years,
            aes(x = year, y = fit),
            color = "red") +
  geom_ribbon(data = future_years,
              aes(x = year, y = fit, ymin = lwr, ymax = upr),
              alpha = 0.3, fill = "red") +
  labs(y = "Heat advisory days",
       y = "Year") +
  theme_minimal()


dat1 %>% 
  filter(yday > 152 & yday < 260) %>% 
  filter(heat_index > 95)

```

## Session Information

```{r Session_Info, echo = F, comment = ""}

# Add session information to help with reproduceability
sessionInfo()


```


